---
title: "gt styling"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    highlight-style: a11y
    link-external-newwindow: true
    embed-resources: true
    grid:
      sidebar-width: 300px
      body-width: 900px
      margin-width: 300px
      gutter-width: 1.5rem
    theme: lux
execute: 
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| include: FALSE

library(tidyverse) ; library(gt)
```


```{r}
#| label: load-data
#| echo: FALSE

df <- countrypops |> 
  filter(country_code_3 %in% c("DZA","BHR","COM","DJI","EGY","IRQ","JOR","KWT",
                               "LBN","LBY","MRT","MAR","OMN","PSE","QAT","SAU",
                               "SOM","SDN","SYR","TUN","ARE","YEM")) |>
  select(country_name, year, population) |>
  filter(year %in% c(2012,2022)) |>
  pivot_wider(
    names_from = "year",
    values_from = "population"
  ) |>
  mutate(government = case_when(
    country_name %in% c("Bahrain","Jordan","Kuwait","Morocco","Oman","Qatar","Saudi Arabia","United Arab Emirates") ~ "Monarchy",
    TRUE ~ "Republic"
  ),
  change = (`2022`-`2012`)/`2012`) |> 
  relocate(government, .after = country_name) |> 
  arrange(desc(change))
```

This document shows some of the ways that you can format and style tables created using the [**`gt`**](https://gt.rstudio.com) R package.

```{r}
#| label: tbl-unstyled
#| tbl-cap: "Unstyled"
#| include: FALSE

tbl <- df |> 
  gt(rowname_col = "country_name",
     groupname_col = "government") |> 
  tab_stubhead(label = "Government type") |>
  cols_label(change = "Change") |> 
  tab_spanner(columns = c(`2012`,`2022`), label = "Year") |>
  fmt_number(columns = c(`2012`,`2022`), decimals = 2, suffixing = TRUE) |> 
  fmt_percent(columns = change, decimals = 1) |> 
  summary_rows(
    columns = c(`2012`, `2022`),
    fns = list(Total = ~ sum(.)), 
    fmt = ~ fmt_number(., decimals = 1, suffixing = TRUE)
  ) |>
  summary_rows(
    groups = everything(),
    columns = "change",
    fns = list(Total = ~ ((sum(`2022`) - sum(`2012`)) / sum(`2012`))),
    fmt = ~ fmt_percent(., decimals = 1)
  ) |> 
  grand_summary_rows(
    columns = c(`2012`, `2022`),
    fns =  list(label = "Grand Total", id = "total", fn = "sum"),
    fmt = ~ fmt_number(., decimals = 2, suffixing = TRUE)
  ) |> 
  grand_summary_rows(
    columns = "change",
    fns = list("Grand Total" = ~ ((sum(`2022`) - sum(`2012`)) / sum(`2012`))),
    fmt = ~ fmt_percent(., decimals = 1)
  ) |> 
  tab_header(title = "Population of countries belonging to the League of Arab States") |> 
  #opt_table_lines("none") |> 
  tab_options(
    table.font.size = px(10),
    data_row.padding = px(2)
    )
```

**`gt`** tables have a clearly articulated structure as shown in the syntax used in @fig-gt.

```{r}
#| label: fig-gt
#| fig-align: left
#| fig-cap: "Source: gt.rstudio.com"
#| echo: FALSE
knitr::include_graphics("https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg")
```

A **`gt`** table can be styled by targeting parts such as the **Spanner** or **Stubhead** by using the corresponding [`helper` functions](https://gt.rstudio.com/reference/index.html#helper-functions). We will generally apply formatting and styles to these table `locations` using the [`tab_style()`](https://gt.rstudio.com/reference/tab_style.html) or [`tab_options()`](https://gt.rstudio.com/reference/tab_options.html) functions.

## Typeface

::: {.panel-tabset}
### Title
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
      style = cell_text(
        font = list(
          google_font(name = "Walter Turncoat"))
        ),
      locations = cells_title(groups = "title")
      )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
      style = cell_text(
        font = list(
          google_font(name = "Walter Turncoat"))
        ),
      locations = cells_title(groups = "title")
      )
```
:::
:::

### Column labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Playfair"), 
      weight = "bold",
      size = px(20)
      ),
    locations = cells_column_labels(everything())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Playfair"), 
      weight = "bold",
      size = px(20)
      ),
    locations = cells_column_labels(everything())
  )
```
:::
:::

### Cells
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Roboto Mono")),
    locations = cells_body(columns = everything())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Roboto Mono")),
    locations = cells_body(columns = everything())
  )
```
:::
:::

### Table
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_font(font = google_font("Source Sans Pro"))
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_font(font = google_font("Source Sans Pro"))
```
:::
:::
:::

## Font

::: {.panel-tabset}
### Stub head label
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  )
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  )
```
:::
:::

### Row group labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_options(row_group.font.weight = "bold")
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(row_group.font.weight = "bold")
```
:::
:::

### Column labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_column_labels(columns = everything())
  )
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_column_labels(columns = everything())
  )
```
:::
:::

### Spanner labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(color =	"#756bb1", weight = "bold"),
    locations = cells_column_spanners()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(color =	"#756bb1", weight = "bold"),
    locations = cells_column_spanners()
  )
```
:::
:::

### Column and spanner labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_options(column_labels.font.weight = "bold")
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(column_labels.font.weight = "bold")
```
:::
:::

### Cells
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = "change")
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = "change")
  )
```
:::
:::

### Summary row
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_summary(),
                     cells_summary())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_summary(),
                     cells_summary())
  )
```
:::
:::

### Grand Summary row
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |>  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_grand_summary(),
                     cells_grand_summary())
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |>  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_grand_summary(),
                     cells_grand_summary())
    )
```
:::
:::
:::

## Alignment
::: {.panel-tabset}
### Stub head label
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center", v_align = "middle"),
    locations = cells_stubhead()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center", v_align = "middle"),
    locations = cells_stubhead()
  )
```
:::
:::

### Column label
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())
  )
```
:::
:::
:::

## Width
::: {.panel-tabset}
### Column
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  cols_width(
    country_name ~ px(180),
    everything() ~ px(60)
)
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  cols_width(
    country_name ~ px(120),
    everything() ~ px(60)
)
```
:::
:::

### Table
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_options(
    table.width = pct(50)
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(
    table.width = pct(50)
  )
```
:::
:::
:::

## Borders
::: {.panel-tabset}
### Stub
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    stub.border.style = "solid",
    stub.border.width = px(2),
    stub.border.color = "#000000"
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    stub.border.style = "solid",
    stub.border.width = px(2),
    stub.border.color = "#000000"
  )
```
:::
:::

### Row group
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
   opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_row_groups()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_row_groups()
  )
```
:::
:::

### Column labels
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = cells_column_labels(columns = everything())
  ) |> 
  tab_style(
    style = cell_borders(sides = "right", style = "hidden"),
    locations = cells_column_spanners("Year")
    ) 
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = cells_column_labels(columns = everything())
  ) |> 
  tab_style(
    style = cell_borders(sides = "right", style = "hidden"),
    locations = cells_column_spanners("Year")
    ) 
```
:::
:::

### Column cells
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = list(
    cells_body(columns = `2022`),
    cells_summary(columns = `2022`))
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = list(
    cells_body(columns = `2022`),
    cells_summary(columns = `2022`))
  )
```
:::
:::

### Specific row cells
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_body(rows = country_name == "United Arab Emirates")
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_body(rows = country_name == "United Arab Emirates")
    )
```
:::
:::

### Table
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    heading.border.lr.style = "hidden",
    heading.border.bottom.style = "solid",
    heading.border.bottom.width = px(2),
    heading.border.bottom.color = "#000000",
    table.border.left.style = "solid",
    table.border.left.width = px(2),
    table.border.left.color = "#000000",
    table.border.right.style = "solid",
    table.border.right.width = px(2),
    table.border.right.color = "#000000",
    table.border.bottom.color = "#000000",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    heading.border.lr.style = "hidden",
    heading.border.bottom.style = "solid",
    heading.border.bottom.width = px(2),
    heading.border.bottom.color = "#000000",
    table.border.left.style = "solid",
    table.border.left.width = px(2),
    table.border.left.color = "#000000",
    table.border.right.style = "solid",
    table.border.right.width = px(2),
    table.border.right.color = "#000000",
    table.border.bottom.color = "#000000",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
  )
```
:::
:::
:::

## Highlighting
::: {.panel-tabset}
### Single value
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
        style = cell_text(color = "red", weight = "bold"),
        locations = cells_body(columns = change, 
                               rows = change == min(change))
        )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
        style = cell_text(color = "red", weight = "bold"),
        locations = cells_body(columns = change, 
                               rows = change == min(change))
        )
```
:::
:::

### Set of values
::: grid
::: g-col-6
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(color = "seagreen", weight = "bold"),
    locations = list(cells_stub(rows = change >= 0.4),
                     cells_body(rows = change >= 0.4))
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(color = "seagreen", weight = "bold"),
    locations = list(cells_stub(rows = change >= 0.4),
                     cells_body(rows = change >= 0.4))
  )
```
:::
:::
:::

## Download button
::: grid
::: g-col-6
```{r}
#| eval: FALSE
library(downloadthis)
tbl |> 
  tab_source_note(
    df |> 
      download_this(
        output_name = "tbl-17",
        output_extension = ".csv",
        csv2 = FALSE,
        button_label = "Download CSV",
        button_type = "default"
      )
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
library(downloadthis)
tbl |> 
  tab_source_note(
    df |> 
      download_this(
        output_name = "tbl-17",
        output_extension = ".csv",
        csv2 = FALSE,
        button_label = "Download CSV",
        button_type = "default"
      )
  )
```
:::
:::