---
title: "gt styling"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    highlight-style: a11y
    link-external-newwindow: true
    embed-resources: true
    grid:
      sidebar-width: 300px
      body-width: 900px
      margin-width: 300px
      gutter-width: 1.5rem
    theme: minty
execute: 
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

This page shows some of the ways that you can style tables created using the [**`gt`**](https://gt.rstudio.com) R package.

A **`gt`** table can be styled using the [`tab_styles()`](https://gt.rstudio.com/reference/tab_style.html), [`tab_options()`](https://gt.rstudio.com/reference/tab_options.html) or [`opt_stylize()`](https://gt.rstudio.com/reference/opt_stylize.html) functions. However, you need to know where to target your styles before you can apply them. 

Each **`gt`** table has a formal structure with clearly defined parts: a table header, stub and stub head, column labels, a table body and a table footer.

```{r}
#| label: fig-gt
#| fig-align: left
#| fig-cap: "Source: gt.rstudio.com"
#| echo: FALSE
knitr::include_graphics("https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg")
```

You can apply styles to part(s) of a **`gt`** table by passing `cell_*()` helper functions to the `style` and `locations` arguments of `tab_style()` For example, if you wanted to highlight the third column in a table you might run:

```{r}
#| eval: FALSE
df |> 
  gt() |> 
  tab_style(
    style = cell_fill(color = "honeydew2"), # <1>
    locations = cells_body(columns = 3) # <2>
  )
```

1. where [`cell_fill()`](https://gt.rstudio.com/reference/cell_fill.html) is the style to apply and 
2. [`cells_body()`](https://gt.rstudio.com/reference/cells_body.html) is the component of the table that you wish to target

We will be applying different styles to a `gt` table that displays a cut of the [`countrypops`](https://gt.rstudio.com/reference/countrypops.html) dataset. The table includes row groupings and summary totals to help illustrate the different components that can be targeted.

```{r}
#| code-fold: TRUE
#| code-summary: "Show the code to create the table"

library(tidyverse) ; library(gt)

tbl <- countrypops |> 
  filter(country_code_3 %in% c("DZA","BHR","COM","DJI","EGY","IRQ","JOR","KWT",
                               "LBN","LBY","MRT","MAR","OMN","PSE","QAT","SAU",
                               "SOM","SDN","SYR","TUN","ARE","YEM")) |>
  select(country_name, year, population) |>
  filter(year %in% c(2012,2022)) |>
  pivot_wider(
    names_from = "year",
    values_from = "population"
  ) |>
  mutate(government = case_when(
    country_name %in% c("Bahrain","Jordan","Kuwait","Morocco","Oman","Qatar","Saudi Arabia","United Arab Emirates") ~ "Monarchy",
    TRUE ~ "Republic"
  ),
  change = (`2022`-`2012`)/`2012`) |> 
  relocate(government, .after = country_name) |> 
  arrange(desc(change)) |> 
  gt(rowname_col = "country_name",
     groupname_col = "government") |> 
  tab_stubhead(label = "Government type") |>
  cols_label(change = "Change") |> 
  tab_spanner(columns = c(`2012`,`2022`), label = "Year") |>
  fmt_number(columns = c(`2012`,`2022`), decimals = 2, suffixing = TRUE) |> 
  fmt_percent(columns = change, decimals = 1) |> 
  summary_rows(
    columns = c(`2012`, `2022`),
    fns = list(Total = ~ sum(.)), 
    fmt = ~ fmt_number(., decimals = 1, suffixing = TRUE)
  ) |>
  summary_rows(
    groups = everything(),
    columns = "change",
    fns = list(Total = ~ ((sum(`2022`) - sum(`2012`)) / sum(`2012`))),
    fmt = ~ fmt_percent(., decimals = 1)
  ) |> 
  grand_summary_rows(
    columns = c(`2012`, `2022`),
    fns =  list(label = "Grand Total", id = "total", fn = "sum"),
    fmt = ~ fmt_number(., decimals = 2, suffixing = TRUE)
  ) |> 
  grand_summary_rows(
    columns = "change",
    fns = list("Grand Total" = ~ ((sum(`2022`) - sum(`2012`)) / sum(`2012`))),
    fmt = ~ fmt_percent(., decimals = 1)
  ) |> 
  tab_header(title = "Population of countries belonging to the League of Arab States") |> 
  tab_options(
    table.font.size = px(10),
    heading.title.font.size = px(12)
    ) |> 
  opt_vertical_padding(scale = 0.3)
```

## Typeface

::: {.panel-tabset}
### Title
::: grid
::: g-col-6
Apply a [Google Font](https://fonts.google.com/specimen/Walter+Turncoat) to the table title. 

```{r}
#| eval: FALSE
tbl |> 
  tab_style(
      style = cell_text(
        font = list(
          google_font(name = "Walter Turncoat"))
        ),
      locations = cells_title(groups = "title")
      )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
      style = cell_text(
        font = list(
          google_font(name = "Walter Turncoat"))
        ),
      locations = cells_title(groups = "title")
      )
```
:::
:::

### Column labels
::: grid
::: g-col-6
Apply a [Google Font](https://fonts.google.com/specimen/Merriweather) to all the column labels and change the size.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Merriweather"), 
      size = px(10)
      ),
    locations = cells_column_labels(everything())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Merriweather"), 
      size = px(10)
      ),
    locations = cells_column_labels(everything())
  )
```
:::
:::

### Cells
::: grid
::: g-col-6
Apply a [Google Font](https://fonts.google.com/specimen/Roboto+Mono) to all the table cells.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Roboto Mono")),
    locations = cells_body(columns = everything())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(
      font = google_font("Roboto Mono")),
    locations = cells_body(columns = everything())
  )
```
:::
:::

### Table
::: grid
::: g-col-6
Apply a [themed set of fonts](https://gt.rstudio.com/reference/system_fonts.html) to the entire table.
```{r}
#| eval: FALSE
tbl |> 
  opt_table_font(stack = "rounded-sans")
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_font(stack = "rounded-sans")
```
:::
:::
:::

## Font

::: {.panel-tabset}
### Stub head label
::: grid
::: g-col-6
Set the stub head label weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  )
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  )
```
:::
:::

### Row group labels
::: grid
::: g-col-6
Set the row group label weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_options(row_group.font.weight = "bold")
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(row_group.font.weight = "bold")
```
:::
:::

### Column labels
::: grid
::: g-col-6
Set the column label weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = everything())
  )
```
:::

::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = everything())
  )
```
:::
:::

### Spanner labels
::: grid
::: g-col-6
Set the spanner label weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  )
```
:::
:::

### Column and spanner labels
::: grid
::: g-col-6
Set the column and spanner label weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_options(column_labels.font.weight = "bold")
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(column_labels.font.weight = "bold")
```
:::
:::

### Cells
::: grid
::: g-col-6
Set the table cell weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = "change")
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = "change")
  )
```
:::
:::

### Summary row
::: grid
::: g-col-6
Set the summary row weight to bold.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_summary(),
                     cells_summary())
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_summary(),
                     cells_summary())
  )
```
:::
:::

### Grand Summary row
::: grid
::: g-col-6
Set the grand summary weight to bold.
```{r}
#| eval: FALSE
tbl |>  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_grand_summary(),
                     cells_grand_summary())
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |>  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_stub_grand_summary(),
                     cells_grand_summary())
    )
```
:::
:::
:::

## Width
::: {.panel-tabset}
### Column
::: grid
::: g-col-6
Adjust the width of the `country_name` column and fix the other columns to the same width.
```{r}
#| eval: FALSE
tbl |> 
  cols_width(
    country_name ~ px(180),
    everything() ~ px(60)
)
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  cols_width(
    country_name ~ px(120),
    everything() ~ px(60)
)
```
:::
:::

### Table
::: grid
::: g-col-6
Reduce the width of the table to 50%.
```{r}
#| eval: FALSE
tbl |> 
  tab_options(
    table.width = pct(50)
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_options(
    table.width = pct(50)
  )
```
:::
:::
:::

## Alignment
::: {.panel-tabset}
### Stub head label
::: grid
::: g-col-6
Align the stub head label to the centre and vertically align it to the middle.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center", v_align = "middle"),
    locations = cells_stubhead()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center", v_align = "middle"),
    locations = cells_stubhead()
  )
```
:::
:::

### Column label
::: grid
::: g-col-6
Centre align all of the column labels
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_column_labels(columns = everything())
  )
```
:::
:::

### Spanner column label
::: grid
::: g-col-6
Left align the column spanner label.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_column_spanners(spanners = "Year")
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_column_spanners(spanners = "Year")
    )
```
:::
:::

### Row names
::: grid
::: g-col-6
Right align the row names
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "right"),
    locations = cells_stub()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = cell_text(align = "right"),
    locations = cells_stub()
  )
```
:::
:::
:::

## Borders
::: {.panel-tabset}
### Stub
::: grid
::: g-col-6
Add a border to the define the end of the stub
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    stub.border.style = "solid",
    stub.border.width = px(2),
    stub.border.color = "#000000"
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    stub.border.style = "solid",
    stub.border.width = px(2),
    stub.border.color = "#000000"
  )
```
:::
:::

### Row group
::: grid
::: g-col-6
Add a horizontal border separating each row group
```{r}
#| eval: FALSE
tbl |> 
   opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_row_groups()
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_row_groups()
  )
```
:::
:::

### Column labels
::: grid
::: g-col-6
Separate each column label with a border
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = cells_column_labels(columns = everything())
  ) |> 
  tab_style(
    style = cell_borders(sides = "right", style = "hidden"),
    locations = cells_column_spanners("Year")
    ) 
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
  style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
  locations = cells_column_labels(columns = everything())
  ) |> 
  tab_style(
    style = cell_borders(sides = "right", style = "hidden"),
    locations = cells_column_spanners("Year")
    ) 
```
:::
:::

### Column cells
::: grid
::: g-col-6
Add a border to the left edge of the `2022` column
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
    locations = list(
      cells_body(columns = `2022`),
      cells_summary(columns = `2022`),
      cells_summary(columns = `2022`))
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "left", color = "#000000", weight = px(2)),
    locations = list(
      cells_body(columns = `2022`),
      cells_summary(columns = `2022`),
      cells_summary(columns = `2022`))
  )
```
:::
:::

### Specific row cells
::: grid
::: g-col-6
Add a horizontal border to a specific row of values
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_body(rows = country_name == "United Arab Emirates")
    )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_style(
    style = cell_borders(sides = "bottom", color = "#000000", weight = px(2)),
    locations = cells_body(rows = country_name == "United Arab Emirates")
    )
```
:::
:::

### Table
::: grid
::: g-col-6
Add an outline to the table
```{r}
#| eval: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    heading.border.lr.style = "hidden",
    heading.border.bottom.style = "solid",
    heading.border.bottom.width = px(2),
    heading.border.bottom.color = "#000000",
    table.border.left.style = "solid",
    table.border.left.width = px(2),
    table.border.left.color = "#000000",
    table.border.right.style = "solid",
    table.border.right.width = px(2),
    table.border.right.color = "#000000",
    table.border.bottom.color = "#000000",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  opt_table_lines("none") |> 
  tab_options(
    heading.border.lr.style = "hidden",
    heading.border.bottom.style = "solid",
    heading.border.bottom.width = px(2),
    heading.border.bottom.color = "#000000",
    table.border.left.style = "solid",
    table.border.left.width = px(2),
    table.border.left.color = "#000000",
    table.border.right.style = "solid",
    table.border.right.width = px(2),
    table.border.right.color = "#000000",
    table.border.bottom.color = "#000000",
    table.border.bottom.style = "solid",
    table.border.bottom.width = px(2),
  )
```
:::
:::
:::

## Highlighting
::: {.panel-tabset}
### Single value
::: grid
::: g-col-6
Highlight the largest percentage change in population between 2012 and 2022.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
        style = cell_text(color = "red", weight = "bold"),
        locations = cells_body(columns = change, 
                               rows = change == max(change))
        )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
        style = cell_text(color = "red", weight = "bold"),
        locations = cells_body(columns = change, 
                               rows = change == max(change))
        )
```
:::
:::

### Set of values
::: grid
::: g-col-6
Highlight rows where the population change was 30% or greater.
```{r}
#| eval: FALSE
tbl |> 
  tab_style(
    style = list(cell_fill(color = "#3182bd", alpha = 0.8),
                 cell_text(color = "white")),
    locations = list(cells_stub(rows = change >= 0.3),
                     cells_body(rows = change >= 0.3))
  )
```
:::
  
::: g-col-6
```{r}
#| echo: FALSE
tbl |> 
  tab_style(
    style = list(cell_fill(color = "#3182bd", alpha = 0.8),
                 cell_text(color = "white")),
    locations = list(cells_stub(rows = change >= 0.3),
                     cells_body(rows = change >= 0.3))
  )
```
:::
:::
:::